Making Sequential Programs Deterministic

### 2019-03-25 design.tex
typesetting

$\infer{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 <: \tau_2}$

@@

$\forall i : i >= 0 \infer{method(\overline{a_i}) : \tau_r}{\tau_r \  method(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$

\subsection{Determinism type hierarchy}\label{type-hierarchy}
The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<@NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<@Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

In this type system, \<@Det> is a subtype of \<@NonDet> denoted as \codeid{@Det <: @NonDet}.

@@

The underlying checker-framework on which we build our determinism checker has support for
polymorphic annotations. When a user writes a polymorphic annotation on a method signature or a type parameter,
it indicates that it could resolve to any type in the type system depending on how it is used.
In the determinism checker, we define a polymorphic annotation \<@PolyDet>.
One of the most common locations for a polymorphic annotation is at a method signature.
Consider the following method declaration annotated with polymorphic annotations:
\begin{verbatim}
@PolyDet int foo(@PolyDet String param1, @PolyDet boolean param2) { }
\end{verbatim}
There are two valid instantiations for this declaration:
\begin{enumerate}
    \item @NonDet int foo(@NonDet String param1, @NonDet boolean param2) { }
    \item @Det int foo(@Det String param1, @Det boolean param2) { }
\end{enumerate}
This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det or @NonDet} type annotations.
%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
 \<@PolyDet> resolves to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, the 
checker framework resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
causing the return type to have the type annotation \<@NonDet>.

\subsection{Handling order nondeterminism}
In addition to the type qualifiers described in~\ref{type-hierarchy}, the determinism checker 
defines the type qualifier \<@OrderNonDet>. This qualifier represents
a collection or an array that will have the same elements in every execution, but in a
possibly different order.  \<@OrderNonDet> may only be written on
collections and arrays.
A collection is any subtype of \<java.util.Collection>,
\<java.util.Iterator>, or \<java.util.Map>, including user-defined types. The relationship
of \<@OrderNonDet> with respect the the other two type qualifiers is \codeid{@Det <: @OrderNonDet <: @NonDet} as shown in Figure~\ref{fig-determinism-hierarchy}.

@@

$\infer[ANY TYPE]{\vdash : \kappa \  \tau}{\kappa : @Det / @NonDet}$

%$\infer[COLLECTION TYPE]{\vdash : \kappa \  \tau}{\kappa : @OrderNonDet & \tau : Collection/Map/Iterator/array}$

$\infer[ORDERNONDET \ COLLECTION]{\vdash : \kappa_c <\kappa_e \ \tau_e> \tau_c}{\kappa_c : @OrderNonDet/@Det & \kappa_e <: \kappa_c & \tau_c <: Collection/Map/Iterator}$

$\infer[ORDERNONDET \ ARRAY]{\vdash : \kappa_c [\kappa_e \ \tau_e] \tau_c}{\kappa_c : @OrderNonDet/@Det & \kappa_e <: \kappa_c }$

$\infer[NONDET \ COLLECTION]{\vdash : \kappa_c <\kappa_e \ \tau_e> \tau_c}{\kappa_c : @NonDet & \kappa_e : @NonDet & \tau_c <: Collection/Map/Iterator}$

$\infer[NONDET \ ARRAY]{\vdash : \kappa_c [\kappa_e \ \tau_e] \tau_c}{\kappa_c : @NonDet & \kappa_e : @NonDet }$

The (determinism) type of a Collection or Iterator must be a supertype or equal to
the type of the type parameter (see Figure~\ref{fig-determinism-collections}).
If the Collection is \<@NonDet>, then the type parameter may not be
\<@Det> or \<@OrderNonDet>. Although such types could exist, they are
disallowed to prevent the following bug:

@@

A collection (\<List, Set, or a Map>) annotated as \<@OrderNonDet> has the following properties:
\begin{enumerate}
    \item The individual elements extracted by iterating over the collection get the type \<@NonDet>.
    \item Accessing properties sich as \<size()> or querying if the collection \<isEmpty()> will return types
    that are annotated as \<@Det> as they do not depend on the iteration order. 
\end{enumerate}

A user defined type may be annotated as \<@OrderNonDet> if and only if it
is a subtype of Collection or Iterator.
For example, if a user defines a type as
\begin{verbatim}
public class TestUserCollection<E> extends ArrayList<E> {...}
\end{verbatim}
Writing \codeid{@OrderNonDet TestUserCollection<@Det Integer>} is valid.\\
Writing \codeid{@OrderNonDet TestUserCollection<@NonDet Integer>} is invalid
because the type of the type parameter (\codeid{@NonDet}) is not a subtype
of the user defined Collection type (\codeid{@OrderNonDet}).

@@

$\infer{\vdash : \kappa \  \tau}{isParam(\tau) / isReturn(\tau) & \kappa : @PolyDet("up") / @PolyDet("down")}$

@@

$\infer{\Gamma \vdash x : @NonDet \ \tau_a}{\Gamma \vdash x : \kappa_a \ \tau_a &  \kappa_a : @OrderNonDet/@NonDet & \kappa_p : @PolyDet("up") & \kappa_p \ \tau_p : declaration \ type & \kappa_a \ \tau_a : resolved \ type}$

@@

$\infer{\Gamma \vdash x : @Det \  \tau_a}{\Gamma \vdash x :  \kappa_a \tau_a & \kappa_a : @OrderNonDet/@Det & \kappa_p : @PolyDet("down") & \kappa_p \ \tau_p : declaration \ type & \kappa_a \ \tau_a : resolved \ type}$

@@

$\infer{}{}$

>>>>

$\infer{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 \sqsubseteq \tau_2}$

@@

$\infer{\mathit{method}(\overline{a_i}) : \tau_r}{\tau_r \  \mathit{method}(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$

\subsection{Determinism type hierarchy}\label{type-hierarchy}
The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

In this type system, \<Det> is a subtype of \<NonDet> denoted as \codeid{Det $\sqsubseteq$ NonDet}.

@@

The underlying checker-framework on which we build our determinism checker has support for
polymorphic annotations. When a user writes a polymorphic annotation on a method signature or a type parameter,
it indicates that it could resolve to any type in the type system depending on how it is used.
In the determinism checker, we define a polymorphic annotation \<PolyDet>.
One of the most common locations for a polymorphic annotation is at a method signature.
Consider the following method declaration annotated with polymorphic annotations:
\begin{verbatim}
@PolyDet int foo(@PolyDet String param1, @PolyDet boolean param2) { }
\end{verbatim}
There are two valid instantiations for this declaration:
\begin{enumerate}
    \item @NonDet int foo(@NonDet String param1, @NonDet boolean param2) { }
    \item @Det int foo(@Det String param1, @Det boolean param2) { }
\end{enumerate}
This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det or @NonDet} type annotations.
%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
 \<@PolyDet> resolves to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, the 
checker framework resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
causing the return type to have the type annotation \<@NonDet>.

\subsection{Handling order nondeterminism}
In addition to the type qualifiers described in~\ref{type-hierarchy}, the determinism checker 
defines the type qualifier \<OrderNonDet>. This qualifier represents
a collection or an array that will have the same elements in every execution, but in a
possibly different order.  \<OrderNonDet> may only be written on
collections and arrays.
A collection is any subtype of \<java.util.Collection>,
\<java.util.Iterator>, or \<java.util.Map>, including user-defined types. The relationship
of \<OrderNonDet> with respect the the other two type qualifiers is \codeid{Det $\sqsubseteq$ OrderNonDet $\sqsubseteq$ NonDet} as shown in Figure~\ref{fig-determinism-hierarchy}.

@@

$\infer[\sc{ANY TYPE}]{\vdash : \kappa \  \beta}{\kappa : \mathit{Det / NonDet}}$

%$\infer[COLLECTION TYPE]{\vdash : \kappa \  \beta}{\kappa : OrderNonDet & \beta : Collection/Map/Iterator/array}$

$\infer[\sc{ORDERNONDET \ COLLECTION}]{\vdash : \kappa_c \langle\kappa_e \ \beta_e\rangle \beta_c}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \mathit{Collection/Map/Iterator}}$

$\infer[\sc{ORDERNONDET \ ARRAY}]{\vdash : \kappa_c [\kappa_e \ \beta_e] \beta_c}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c }$

$\infer[\sc{NONDET \ COLLECTION}]{\vdash : \kappa_c \langle\kappa_e \ \beta_e\rangle \beta_c}{\kappa_c : \mathit{NonDet} & \kappa_e : \mathit{NonDet} & \beta_c \sqsubseteq \mathit{Collection/Map/Iterator}}$

$\infer[\sc{NONDET \ ARRAY}]{\vdash : \kappa_c [\kappa_e \ \beta_e] \beta_c}{\kappa_c : \mathit{NonDet} & \kappa_e : \mathit{NonDet} }$

The (determinism) type of a Collection or Iterator must be a supertype or equal to
the type of the type parameter (see Figure~\ref{fig-determinism-collections}).
If the Collection is \<NonDet>, then the type parameter may not be
\<Det> or \<OrderNonDet>. Although such types could exist, they are
disallowed to prevent the following bug:

@@

A collection (\<List, Set, or a Map>) annotated as \<@OrderNonDet> has the following properties:
\begin{enumerate}
    \item The individual elements extracted by iterating over the collection get the type \<NonDet>.
    \item Accessing properties sich as \<size()> or querying if the collection \<isEmpty()> will return types
    that are annotated as \<Det> as they do not depend on the iteration order. 
\end{enumerate}

A user defined type may be annotated as \<OrderNonDet> if and only if it
is a subtype of Collection or Iterator.
For example, if a user defines a type as
\begin{verbatim}
public class TestUserCollection<E> extends ArrayList<E> {...}
\end{verbatim}
Writing \codeid{@OrderNonDet TestUserCollection<@Det Integer>} is valid.\\
Writing \codeid{@OrderNonDet TestUserCollection<@NonDet Integer>} is invalid
because the type of the type parameter (\codeid{@NonDet}) is not a subtype
of the user defined Collection type (\codeid{@OrderNonDet}).

@@

$\infer{\vdash : \kappa \  \beta}{isParam(\beta) / isReturn(\beta) & \kappa : @PolyDet("up") / @PolyDet("down")}$

@@

$\infer{\Gamma \vdash x : @NonDet \ \beta_a}{\Gamma \vdash x : \kappa_a \ \beta_a &  \kappa_a : @OrderNonDet/@NonDet & \kappa_p : @PolyDet("up") & \kappa_p \ \beta_p : declaration \ type & \kappa_a \ \beta_a : resolved \ type}$

@@

$\infer{\Gamma \vdash x : @Det \  \beta_a}{\Gamma \vdash x :  \kappa_a \beta_a & \kappa_a : @OrderNonDet/@Det & \kappa_p : @PolyDet("down") & \kappa_p \ \beta_p : declaration \ type & \kappa_a \ \beta_a : resolved \ type}$

@@
